\part{Calculus III}


\apexchapter{Vectors}{chapter:vectors}

This chapter introduces a new mathematical object, the \sword{vector}. Defined in \autoref{sec:vector_intro}, we will see that vectors provide a powerful language for describing quantities that have magnitude and direction aspects. A simple example of such a quantity is force: when applying a force, one is generally interested in how much force is applied (i.e., the magnitude of the force) and the direction in which the force was applied. Vectors will play an important role in many of the subsequent chapters in this text. 

This chapter begins with moving our mathematics out of the plane and into ``space.'' That is, we begin to think mathematically not only in two dimensions, but in three. With this foundation, we can explore vectors both in the plane and in space. 

\input{text/10_Space_Intro}
\input{text/10_Vector_Introduction}
\input{text/10_Dot_Product}
\input{text/10_Cross_Product}
\input{text/10_Lines}
\input{text/10_Planes}
\input{text/10_Other_Systems}


\apexchapter{Vector Valued Functions}{chap:vvf}

In the previous chapter, we learned about vectors and were introduced to the power of vectors within mathematics. In this chapter, we'll build on this foundation to define functions whose input is a real number and whose output is a vector. We'll see how to graph these functions and apply calculus techniques to analyze their behavior. Most importantly, we'll see \textit{why} we are interested in doing this: we'll see beautiful applications to the study of moving objects.

\input{text/11_Vector_Functions_Intro}
\input{text/11_Vector_Functions_Calc}
\input{text/11_Vector_Functions_Motion}
\input{text/11_Vector_Tangent_Normal}
\input{text/11_Arc_Length_Parameter_Curvature}


\apexchapter{Functions of Several Variables}{chap:multi}

A function of the form $y=f(x)$ is a function of a single variable; given a value of $x$, we can find a value $y$. Even the vector--valued functions of \autoref{chap:vvf} are single--variable functions; the input is a single variable though the output is a vector.

There are many situations where a desired quantity is a function of two or more variables. For instance, wind chill is measured by knowing the temperature and wind speed; the volume of a gas can be computed knowing the pressure and temperature of the gas; to compute a baseball player's batting average, one needs to know the number of hits and the number of at--bats. 

This chapter studies \textbf{multivariable} functions, that is, functions with more than one input.

\input{text/12_Multivariable_Intro}
\input{text/12_Multivariable_Limit}
\input{text/12_Partial_Derivatives}
\input{text/12_Total_Differential}
\input{text/12_Multivariable_Chain_Rule}
\input{text/12_Directional_Derivatives}
\input{text/12_Tangent_Planes_Lines}
\input{text/12_Extrema}
\input{text/12_Lagrange}


\apexchapter{Multiple Integration}{chapter:mult_int}

The previous chapter introduced multivariable functions and we applied concepts of differential calculus to these functions. We learned how we can view a function of two variables as a surface in space, and learned how partial derivatives convey  information about how the surface is changing in any direction.

In this chapter we apply techniques of integral calculus to multivariable functions. In \autoref{chapter:integration} we learned how the definite integral of a single variable function gave us ``area under the curve." In this chapter we will see that integration applied to a multivariable function gives us ``volume under a surface." And just as we learned applications of integration beyond finding areas, we will find applications of integration in this chapter beyond finding volume.

\input{text/13_Iterated_Integrals}
\input{text/13_Double_Integrals_Volume}
\input{text/13_Double_Integrals_Polar}
\input{text/13_Center_of_Mass}
\input{text/13_Surface_Area}
\input{text/13_Triple_Integration}
\input{text/13_Jacobian}


\apexchapter{Line and Surface Integrals}{chapter:vect_calc}

% todo Chapter 15 needs an introduction
\input{text/14_Line_Integrals}
\input{text/14_Line_Integrals_Props}
\input{text/14_Greens}
\input{text/14_Div_Thm}
\input{text/14_Stokes}
\input{text/14_Div_Grad_Curl}

\cleardoublepage
