\section*{13.2 \ \  Possible Additional Material}


(The Squeeze Theorem.)

(Maybe as a boxed formula/theorem right before or right after Example 5.)

Let $S$ be an open set containing $(x_0, y_0)$.  Suppose $f(x, y)$, $g(x, y)$, and $h(x, y)$ are defined on $S$ except possibly at $(x_0, y_0)$ and both
\[
\lim_{(x,y) \rightarrow (x_0, y_0)} g(x, y) = L \ \ \text{and} \ \ \lim_{(x,y) \rightarrow (x_0, y_0)} h(x, y) = L.
\]
If $g(x, y) \leq f(x, y) \leq h(x, y)$ for all $(x, y) \in S$ except possibly at $(x_0, y_0)$ then
\[
\lim_{(x,y) \rightarrow (x_0, y_0)} f(x, y) = L.
\]

*****

(A Solution 2 for Example 5.)

We have 
\[
0 \leq \frac{5x^2y^2}{x^2 + y^2} \leq \frac{5x^2y^2 + 5y^4}{x^2 + y^2}
= \frac{5y^2(x^2+y^2)}{x^2 + y^2} = 5y^2.
\]
Since $0 \rightarrow 0$ and $5y^2 \rightarrow 0$ as $(x, y) \rightarrow (0,0)$ we have 
\[
\lim_{(x,y) \rightarrow (0, 0)} \frac{5x^2y^2}{x^2 + y^2} = 0
\]
by the Squeeze Theorem.

*****

(A Solution 3 for Example 5.)

If we set $x = r \cos \theta$ and $y = r \sin \theta$ we have $(x, y) \rightarrow 0$ as $r \rightarrow 0$.  Now
\[
\frac{5x^2y^2}{x^2 + y^2} = \frac{5r^4 \cos^2 \theta \sin^2 \theta}{r^2(\cos^2 \theta + \sin^2 \theta)} = 5r^2 \cos^2 \theta \sin^2 \theta.
\]
Thus 
\[
0 \leq \frac{5x^2y^2}{x^2 + y^2} = 5r^2 \cos^2 \theta \sin^2 \theta \leq 5r^2.
\]
Since $5r^2 \rightarrow 0$ as $r \rightarrow 0$ we have again
\[
\lim_{(x,y) \rightarrow (0, 0)} \frac{5x^2y^2}{x^2 + y^2} = 0
\]
by the Squeeze Theorem.

*****

(After Definition 84.)

It follows that if $f$ is a continuous function of one variable then $f(x)$ (or $f(y)$) is a continuous function of the variables $(x, y)$.

*****

\section*{13.2 \ \ More Problems}

\ 


\begin{enumerate}
\item  Use the Squeeze Theorem to show
\[
\lim_{(x,y) \rightarrow (0,0)} \sin(xy) \cos\left(\frac{1}{x^2+y^2}\right) = 0.
\]  
[Solution/hint:  Consider $-|\sin(xy)| \leq \sin(xy) \cos\left(\frac{1}{x^2+y^2}\right) \leq |\sin(xy)|$.]  \\

\item  Let $f(x, y) = \dfrac{x^2y}{x^4 + y^2}$.
\begin{enumerate}
\item  Show that $f(x, y) \rightarrow 0$ along every line through the origin. \\


\item  Despite this, show that $\lim_{(x, y) \rightarrow (0, 0)} f(x, y)$ does not exist.
\end{enumerate}
[Solution/hint:  For the second part consider the path $y = x^2$.]  \\


\item  Show that 
\[
\lim_{(x,y) \rightarrow (0,0)} \frac{x^4 - 9y^4}{x^2 + 3y^2} = 0.
\]
[Solution/hint:  $x^4 - 9y^4 = (x^2 + 3y^2)(x^2 - 3y^2)$.]  \\

\item  Determine the domain of the following functions.  Explain why the function is continuous on its domain.
\begin{enumerate}
\item  $f(x, y) = x^2 + 3xy + y^2$

[Solution:  The domain is all of $\mathbb{R}^2$.  Appeal to Theorems 104 and 105.]  \\

\item  $f(x, y) = \dfrac{x^2 + y + 1}{x - 2y + 3}$

[Solution:  The domain is all $(x, y)$ in $\mathbb{R}^2$ with $x - 2y + 3 \neq 0$.  This is all points in the plane except those on the line $y = \frac{1}{2}x - \frac{3}{2}$.  Appeal to Theorems 104 and 105.]  \\

\item  $f(x, y) = \dfrac{1}{\sqrt{1 - x^2 - y^2}}$

[Solution:  The domain is all $(x, y)$ in $\mathbb{R}^2$ with $x^2 + y^2 < 1$.  This is the open disk of radius 1 centered at the origin.  Appeal to Theorems 104 and 105.]  \\

\item  $f(x, y) = \ln(xy)$

[Solution:  The domain is all $(x, y)$ in $\mathbb{R}^2$ with either $x, y > 0$ or $x, y < 0$.  This is the first and third quadrants of the $xy$-plane without the axes.  Appeal to Theorems 104 and 105.]  \\

\end{enumerate}

\end{enumerate}