\documentclass[10pt]{article}


\usepackage{ifthen}

\usepackage{lipsum}
\usepackage{pgfplots}

\usepackage{eso-pic,calc}
\usepackage[font=small]{caption}
\usepgfplotslibrary{external}
\usetikzlibrary{calc}
\usetikzlibrary{shadings}
\usepackage{tikz}
\usetikzlibrary{positioning,chains,fit,shapes,calc,arrows,patterns}
\usepackage{tkz-graph}
\usetikzlibrary{arrows, petri, topaths}
\usepackage{tkz-berge}
\usepackage[all]{xy}
\usepackage{textcomp}
\usepackage[h]{esvect}
\usepackage[normalem]{ulem}

\pgfplotsset{compat=1.8}
\usepackage{amssymb}

\usepackage{amsmath}

\newcommand{\ds}{\displaystyle}

\begin{document} 

%%%%% Insert the following text right before example 128.

How would we measure total distance traveled? We have to consider the intervals when $v(t)\geq 0$ and when $v(t)\leq 0$. Therefore, $$\text{total distance traveled}=\int_a^b |v(t)| ~dt$$.

Change the title of Example 128 to be Finding displacement and total distance traveled

Change the last sentence of the directions to Example 128 to read:  Find and interpret, 1. $\ds \int_0^1 v(t) ~dt$ and 2. $\ds\int_ 0^1 |v(t)| ~dt$.

Solution to part 1. reads:

{Using the Fundamental Theorem of Calculus, we have 
\begin{align*}
\int_0^1 v(t)\ dt &= \int_0^1 (-32t+20)\ dt \\
			&= -16t^2 + 20t\Big|_0^1 \\
			&= 4 \text{\ ft}.
\end{align*}
Thus if a ball is thrown straight up into the air with velocity $v(t) = -32t+20$, the height of the ball, 1 second later, will be 4 feet above the initial height. We will see in part 2. that the \textit{distance traveled} is much farther. It has gone up to its peak and is falling down, but the difference between its height at $t=0$ and $t=1$ is 4 ft.}\\

Solution to part 2. reads: Here we are trying to find the total distance traveled by the ball. We must first consider where $v(t)>0$ and $v(t)<0$.  
\begin{align*}
v(t)=-32t+20&=0\\
-32t&=-20\\
t&=\frac{5}{8}
\end{align*}

$v(t)>0$ for $t<\frac{5}{8}$ and $v(t)<0$ for $t>\frac{5}{8}$ so we have 
\begin{align*}
\int_0^1 |v(t)|~dt&=\int_0^{5/8} v(t)~dt + \int_{5/8}^1 -v(t)~dt\\
&=\int_0^{5/8} -32t+20~dt + \int_{5/8}^1 32t-20 ~dt\\
&=\frac{34}{4}=8.5 \text{\ ft}.
\end{align*}

Add the following two exercises after 28.

$\ds \int_0^2 |x^2-1|~dx$   Solution: 2

$\ds \int_0^3|1-2x|~dx$  Solution:  $\frac{7}{2}$

Change directions for 40-44 to read:  ..... Find a) the displacement of the object over the given time interval and b) the total distance traveled by the object over the given time interval. \\


49-52 should be moved to the Area between curves section.

\end{document}