\printproblems

\exerciseset{In Exercises}{, find the (a) cylindrical and (b) spherical coordinates of the point whose Cartesian coordinates are given.}{

\exercise{$(2,2\sqrt3,-1)$}{(a) $(4,\frac\pi3,-1)$; (b) $(\sqrt{17},\frac\pi3,\cos^{-1}\frac{-1}{\sqrt{17}})$}

\exercise{$(-5,5,6)$}{(a) $(5\sqrt2,\frac{3\pi}4,6)$; (b) $(\sqrt{86},\frac{3\pi}4,\cos^{-1}\frac6{\sqrt{86}})$}

\exercise{$(\sqrt{21},-\sqrt{7},0)$}{(a) $(2\sqrt{7},\frac{11\pi}{6},0)$; (b) $(2\sqrt{7},\frac{11\pi}{6},\frac{\pi}{2})$}

\exercise{$(0,\sqrt{2},2)$}{(a) $(\sqrt2,\frac\pi2,2)$; (b) $(\sqrt6,\frac\pi2,\cos^{-1}\frac{\sqrt6}3)$}

}

\exerciseset{In Exercises}{, find the Cartesian coordinates of the point whose cylindrical coordinates are given.}{

\exercise{$(1, \frac{\pi}{3}, -2)$}{$(\frac12,\frac{\sqrt3}2,-2)$}

\exercise{$(2, \frac{5\pi}{6}, 3)$}{$(-\sqrt3,-1,3)$}

\exercise{$(\frac{1}{2}, \frac{\pi}{2}, 1)$}{$(0,\frac12,1)$}

\exercise{$(6, \frac{5\pi}{3}, \sqrt{7})$}{$(3,-3\sqrt3,\sqrt7)$}

}

\exerciseset{In Exercises}{, find the Cartesian coordinates of the point whose spherical coordinates are given.}{

\exercise{$(4,\frac\pi3,\frac\pi4)$}{$(\sqrt2,\sqrt6,2\sqrt2)$}

\exercise{$(1, 0, \frac{\pi}{2})$}{$(1,0,0)$}

\exercise{$(3,\frac{3\pi}2,\frac{5\pi}6)$}{$(0,-\frac32,-\frac{3\sqrt3}2)$}

\exercise{$(2,\frac{7\pi}6,\frac{3\pi}4)$}{$(\frac{\sqrt6}2,\frac{\sqrt2}2,\sqrt2)$}

}

\exerciseset{In Exercises}{, describe the set given by Cartesian inequalities in cylindrical.}{

\exercise{$x^2 + y^2 \leq 3$}{$r\le\sqrt3$}

\exercise{$x^2 + y^2 = 1$ and $z \geq 1$}{$r=1$ and $z\ge1$}

\exercise{$x^2 + y^2 + z^2 = 16$ and $y \geq 0$}{$r^2+z^2=16$ and $0\le\theta\le\pi$}

}

\exerciseset{In Exercises}{, describe the set given by Cartesian inequalities in spherical.}{

\exercise{$x^2 + y^2 + z^2 \leq 3$}{$\rho\le\sqrt3$}

\exercise{$x^2 + y^2 + z^2 = 1$ and $z \leq 0$}{$\rho=1$ and $\phi\ge\frac\pi2$}

\exercise{$x^2 + y^2 + z^2 \geq 4$, $x \geq 0$, and $y \geq 0$}{$\rho\ge4$ and $0\le\theta\le\frac\pi2$}

}

\exerciseset{In Exercises}{, write the given equation or region in (a) cylindrical and (b) spherical coordinates.}{

\exercise{$x^2 + y^2 + z^2 = 25$}{(a) $r^2 + z^2 = 25$; (b) $\rho = 5$}

\exercise{$x^2 + y^2 = 2y$}{(a) $r=2\sin\theta$; (b) $\rho=2\sin\theta\csc\phi$}

\exercise{$x^2 + y^2 + 9z^2 = 36$}{(a) $r^2 + 9z^2 = 36$; (b) $\rho^2 ( 1 + 8 \cos^2 \phi ) = 36$}

\exercise{$x \geq 0$, $y \geq 0$, and $z \geq 0$}{(a) $0\le\theta\le\frac\pi2$ and $z\ge0$; (b) $0\le\theta,\phi\le\frac\pi2$}

\exercise{$z \leq 0$}{(a) $z\le0$; (b) $\frac\pi2\le\phi\le\pi$}

}

\exercise{Describe the intersection of the surfaces whose equations in spherical coordinates are $\theta = \frac{\pi}{2}$ and $\phi = \frac{\pi}{4}$.}{The semi-infinite ray starting at the origin and going through $(0,1,1)$.}

\exercise{Consider a solid sphere of radius $5$ centered at the origin.  Remove the solid circular cylinder of radius $2$ centered along the $z$-axis.  Describe this region with cylindrical inequalities.}{$r^2+z^2\le25$ and $r\ge2$}

\exercise{Describe the region below the top of the sphere of radius $R$ centered at the origin and above the cone $z = \sqrt{x^2 + y^2}$ in spherical coordinates.}{$\rho\le R$ and $\phi\le\frac\pi4$}

\exercise{Give an equation for the sphere of radius $2$ with center at the origin in 
\begin{enumerate}
\item  Cartesian 
\item  cylindrical 
\item  spherical
\end{enumerate}}{(a) $x^2+y^2+z^2=4$; (b) $r^2+z^2=4$; (c) $\rho=2$}

\exercise{Find the center and radius of the sphere given by the spherical equation
\[
\rho = 4 \sin \phi \cos \theta + 6 \sin \phi \sin \theta - 2 \cos \phi.
\]}{center: $(2,3,-1)$, radius: $\sqrt{14}$}
% \rho^2 = 4\rho\sin\phi\cos\theta + 6\rho\sin\phi\sin\theta - 2\rho\cos\phi
% x^2+y^2+z^2 = 4x + 6y - 2z
% x^2-4x+4-4 + y^2-6y+9-9 z^2+2z+1-1 = 0
% (x-2)^2 + (y-3)^2 + (z+1)^2 = 4+9+1 = 14

% todo the solution to 11.7#29, 31-33
\exercise{Show that for $a \ne 0$, the equation $\rho = 2a \sin \phi \, \cos \theta$ in spherical coordinates describes a sphere centered at $(a,0,0)$ with radius $\abs{a}$.}{}

\exercise{Let $P = (a,\theta,\phi)$ be a point in spherical coordinates, with $a > 0$ and $0 < \phi < \pi$. Then $P$ lies on the sphere $\rho = a$. Since $0 < \phi < \pi$, the line segment from the origin to $P$ can be extended to intersect the cylinder given by $r = a$ (in cylindrical coordinates). Find the cylindrical coordinates of that point of intersection.}{$(a,\theta,a \cot \phi )$}

\exercise{Let $P_1$ and $P_2$ be points whose spherical coordinates are $( \rho_1,\theta_1,\phi_1 )$ and $( \rho_2,\theta_2,\phi_2 )$, respectively. Let $\vecv_1$ be the vector from the origin to $P_1$, and let $\vecv_2$ be the vector from the origin to $P_2$. For the angle $\gamma$ between $\vecv_1$ and $\vecv_2$, show that
  \[
   \cos\gamma=\cos\phi_1\cos\phi_2+\sin\phi_1\sin\phi_2\cos(\theta_2-\theta_1).
  \]
  This formula is used in electrodynamics to prove the addition theorem for spherical harmonics, which provides a general expression for the electrostatic potential at a point due to a unit charge.% See pp.\ 100-102 in \cite{jac}.
}{}

\exercise{Show that the distance $d$ between the points $P_1$ and $P_2$ with cylindrical coordinates $( r_1,\theta_1,z_1 )$ and $( r_2,\theta_2,z_2 )$, respectively, is
  \[
   d = \sqrt{r_1^2 + r_2^2 - 2 r_1\,r_2 \cos (\, \theta_2 - \theta_1 \,)
   + ( z_2 - z_1 )^2} \, .
  \]}{Hint: Use the distance formula for Cartesian coordinates.}

\exercise{Show that the distance $d$ between the points $P_1$ and $P_2$ with spherical coordinates $( \rho_1,\theta_1,\phi_1 )$ and $( \rho_2,\theta_2,\phi_2 )$, respectively, is
  \[
   d = \sqrt{\rho_1^2 + \rho_2^2 - 2 \rho_1\,\rho_2 [ \sin \phi_1 \,
    \sin \phi_2 \,\cos ( \, \theta_2 - \theta_1 \, ) +
    \cos \phi_1 \, \cos \phi_2 ]} \, .
  \]}{}
