\documentclass[openany]{book}
\usepackage{hyperref}
\usepackage{bookmark}
\begin{document}

\tableofcontents

\chapter{A}
\section{a}
\pdfbookmark[chapter]{B}{chapC}
\hypersetup{bookmarksdepth=-2}
\section{b}
\chapter{C}\label{chapC}
\hypersetup{bookmarksdepth=1}%back to tocdepth
\section{c}
\chapter{D}
\section{d}

\end{document}

\documentclass[openany]{book}

\usepackage[bookmarksnumbered]{hyperref}
\usepackage{bookmark}

\begin{document}

\tableofcontents

\clearpage

\stepcounter{chapter}\stepcounter{chapter}
\pdfbookmark[chapter]{2 [bookmark] Chapter Title}{chapterlabel}
\addtocounter{section}{-1}
\bgroup
\renewcommand{\addcontentsline}[3]{}
\section{pdf only Section Title}
\egroup
\pdfbookmark[section]{2.0 Section Title}{prereq}
\label{prereq}

This is a prerequisite section.

\hypersetup{bookmarksdepth=-2}
\chapter*{[pdf*] Chapter Title}\label{chapterlabel}
\addcontentsline{toc}{chapter}{\texorpdfstring{\string\numberline{2}}{2 }[pdfacl] Chapter Title}
\hypersetup{bookmarksdepth=2}
\pdfbookmark[section]{Chapter Introduction}{chapterlabel}

This chapter is about stuff.

\section{Displayed Section Title}

This is a section.

\end{document}

Can I have tableofcontents chapter entries not appear in pdf bookmarks?

I'd like to have my table of contents and pdf bookmarks be slightly different.  I can monkey around with counters, \addcontentsline, and \pdfbookmark to mostly achieve what I want.  The only thing I'm having trouble with is have something appear in the table of contents that isn't in the pdf bookmarks.

[This answer](https://tex.stackexchange.com/q/13502/107497) led me to using bookmarksdepth, which does result in the indicated items appearing in the table of contents but not the pdf bookmarks.  Unfortunately, the chapter A bookmark also disappears.  Is there a way to have some entries appear in the table of contents, but not the pdf bookmarks, without having other bookmarks also disappear?

%%%%%%%%%%%%%%%%%%%%%%%%%%%%


For various chapters in my book, I'd like to have a prerequisite section 0.  In the text of the book (lets say for chapter 2), this would be ordered:
2.0 Prerequisite section
2 Chapter Title
2.1 First section
In the \tableofcontents, the prerequisite section should not appear.  I can make all of this happen by fiddling with the appropriate counters and \addcontentsline.

For navigability, however, I would like the section to appear in the pdf bookmarks.  I can't just use \pdfbookmark in the order I just had, because then the prerequisite section would appear under the wrong chapter.  So I'm content with the pdf bookmark order
2 Chapter Title
2.0 Prerequisite section
Chapter Introduction
2.1 First section
The closest I've been able to come is:

(Text appearing in `[]` is to help me determine where various titles came from - it wouldn't be in the document.)

Unfortunately, the \addcontentsline also adds that entry to the pdf bookmarks, so that ``2 Chapter Title'' appears twice.  Is it possible to manually add the [pdfacl] entry to the \tableofcontents without that entry showing up in the pdf bookmarks?

