\printconcepts

\exercise{In your own words, describe what it means for a function to be continuous.}{Answers will vary.}

\exercise{In your own words, describe what the Intermediate Value Theorem states.}{Answers will vary.}

\exercise{What is a ``root'' of a function?}{A root of a function $f$ is a value $c$ such that $f(c)=0$.}

\exercise{Given functions $f$ and $g$ on an interval $I$, how can the Bisection Method be used to find a value $c$ where $f(c) = g(c)$?}{Consider the function $h(x) = g(x) - f(x)$, and use the Bisection Method to find a root of $h$.}

\exercise{T/F:	If $f$ is defined on an open interval containing $c$, and $\ds \lim_{x\to c}f(x)$ exists, then $f$ is continuous at $c$.}{F}

\exercise{T/F: If $f$ is continuous at $c$, then $\ds \lim_{x\to c}f(x)$ exists.}{T}

\exercise{T/F: If $f$ is continuous at $c$, then $\ds \lim_{x\to c^+}f(x) = f(c)$.}{T}

\exercise{T/F: If $f$ is continuous on $[a,b]$, then $\ds\lim_{x\to a^-}f(x) = f(a)$.}{F}

\exercise{T/F: If $f$ is continuous on $[0,1)$ and $[1,2)$, then $f$ is continuous on $[0,2)$.}{F}

\exercise{T/F: The sum of continuous functions is also continuous.}{T}

\printproblems

\input{exercises/01-05-exset-01}

%\questioncolumnbreak

\input{exercises/01-05-exset-02}

\input{exercises/01-05-exset-03}

\exercise{Let $\ds f(x) = \begin{cases}
x^2-1&x < 3 \\
x+5&x\geq 3
\end{cases}$.\\
Is $f$ continuous everywhere?}{Yes. The only ``questionable'' place is at $x=3$, but the left and right limits agree.}

\exercise{Let $\ds f(x) = \begin{cases}
x\sin(\frac1x)&x \ne 0 \\
0&x=0
\end{cases}$.\\
Is $f$ continuous everywhere?}{Yes. The only ``questionable'' place is at $x=0$, but the Squeeze Theorem shows that the limits agree.}

\exercisesetinstructions[Exercises]{ test your understanding of the Intermediate Value Theorem.}

\exercise{Let $f$ be continuous on $[1,5]$ where $f(1) = -2$ and $f(5) = -10$. Does a value $1<c<5$ exist such that $f(c) = -9$? Why/why not?}{Yes, by the Intermediate Value Theorem.}

\exercise{Let $g$ be continuous on $[-3,7]$ where $g(0) = 0$ and $g(2) = 25$. Does a value $-3<c<7$ exist such that $g(c) = 15$? Why/why not?}{Yes, by the Intermediate Value Theorem. In fact, we can be more specific and state such a value $c$ exists in $(0,2)$, not just in $(-3,7)$.}

\exercise{Let $f$ be continuous on $[-1,1]$ where $f(-1) = -10$ and $f(1) = 10$. Does a value $-1<c<1$ exist such that $f(c) = 11$? Why/why not?}{We cannot say; the Intermediate Value Theorem only applies to function values between $-10$ and 10; as 11 is outside this range, we do not know.}

\exercise{Let $h$ be a function on $[-1,1]$ where $h(-1) = -10$ and $h(1) = 10$. Does a value $-1<c<1$ exist such that $h(c) = 0$? Why/why not?}{We cannot say; the Intermediate Value Theorem only applies to continuous functions. As we do know know if $h$ is continuous, we cannot say.}

\exercisesetend

\input{exercises/01-05-exset-05}

\input{exercises/01-05-exset-06}

\input{exercises/01-05-exset-04}

\printreview

\exercise{Let $\ds f(x)= \begin{cases}
x^2-5 & x<5 \\
5x & x \geq 5
\end{cases}$. Find
\begin{multicols}{2}
\begin{enumerate}
\item		$\ds \lim_{x\to 5^-} f(x)$
\item		$\ds \lim_{x\to 5^+} f(x)$
\item		$\ds \lim_{x\to 5} f(x)$
\item		$f(5)$\end{enumerate}
\end{multicols}}{\mbox{}\\[-2\baselineskip]\parbox[t]{\linewidth}{\begin{enumerate}
\item		20
\item		25
\item		Limit does not exist
\item		25
\end{enumerate}}}

\exercise{Numerically approximate the following limits:
\begin{enumerate}
\item	$\ds \lim_{x\to -4/5^+} \frac{x^2-8.2 x-7.2}{x^2+5.8 x+4}$
\item	$\ds \lim_{x\to -4/5^-} \frac{x^2-8.2 x-7.2}{x^2+5.8 x+4}$
\end{enumerate}}{\tagpdfsetup{table/header-rows={1}}\begin{tabular}{cc}
$x$ & $f(x)$ \\ \hline 
$-0.81\phantom{0} $& $-2.34129$ \\
$ -0.801$ & $-2.33413$ \\
$ -0.79\phantom{0} $& $-2.32542 $\\
$ -0.799$ & $-2.33254$
\end{tabular}

The top two lines give an approximation of the limit from the left: $-2.33$. The bottom two lines give an approximation from the right: $-2.33$ as well.}

\exercise{Give an example of function $f(x)$ for which $\ds \lim_{x\to 0} f(x)$ does not exist.}{Answers will vary.}
