\printconcepts

\exercise{Explain in your own words, without using $\epsilon$-$\delta$ formality, why $\ds \lim_{x\to c} b = b$.}{Answers will vary.}

\exercise{Explain in your own words, without using $\epsilon$-$\delta$ formality, why $\ds \lim_{x\to c} x = c$.}{Answers will vary.}

\exercise{What does the text mean when it says that certain functions' ``behavior is `nice' in terms of limits''? What, in particular, is ``nice''?}{Answers will vary.}

\exercise{Sketch a graph that visually demonstrates the Squeeze Theorem.}{Answers will vary.}

\exercise{You are given the following information: 
	\begin{enumext}[start=1]
	\item		$\ds \lim_{x\to 1} f(x) = 0$
	\item		$\ds \lim_{x\to 1} g(x) = 0$
	\item		$\ds \lim_{x\to 1} f(x)/g(x) = 2$
	\end{enumext}
What can be said about the relative sizes of $f(x)$ and $g(x)$ as $x$ approaches 1?}{As $x$ is near 1, both $f$ and $g$ are near 0, but $f$ is approximately twice the size of $g$. (I.e., $f(x) \approx 2g(x)$.)}

\exercise{T/F: $\ds \lim_{x\to1}\ln x = 0$. Use a theorem to defend your answer.}
{T; by \autoref{thm:lim_continuous}, $\lim_{x\to1}\ln x = \ln 1 =0$.}

\printproblems

\input{exercises/01-03-exset-02}

\input{exercises/01-03-exset-03}

\input{exercises/01-03-exset-04}

\input{exercises/01-03-exset-06}

\input{exercises/01-03-exset-05}

%\exercise{$\ds \lim_{x\to0} x\sin\left(\frac{1}{x}\right)$}{$0$}

%\exercise{$\ds \lim_{x\to0} \sin x\cos\left(\frac{1}{x^2}\right)$}{$0$}

%\exercise{\autoref{thm:special_limits} states that $\ds \lim_{x\to0} \frac{\cos x-1}{x}=0$. Work through the following steps to establish this result.
%\begin{enumext}[start=1]
%\item	Show why $\ds \frac{\cos x-1}{x} = \frac{-\sin^2 x}{x(\cos x+1)}$. Hint: multiply $\ds \frac{\cos x-1}{x}$ by 1, written in special form.
%\item	Evaluate $\ds \lim_{x\to 0} \frac{-\sin^2 x}{x(\cos x+1)}$ by breaking the expression $\ds \frac{-\sin^2 x}{x(\cos x+1)}$ into a product of two functions and using \autoref{thm:special_limits}.
%\item		How do the above steps establish that $\ds \lim_{x\to0} \frac{\cos x-1}{x}=0$?
%\end{enumext}}{$0$}

\exercise{Let $f(x) = 0$ and $\ds g(x) = \frac xx$.
\begin{enumext}[start=1]
	\item Show why $\ds \lim_{x\to2} f(x)=0$.
	\item Show why $\ds \lim_{x\to0} g(x) = 1$.
	\item Show why $\ds \lim_{x\to2} g\big(f(x)\big)$ does not exist.
	\item Show why the answer to part (c) does not violate \autoref{thm:limit_composition}.
\end{enumext}}
{\begin{enumext}[start=1]
	\item Apply Part 1 of \autoref{thm:limit_algebra}.
	\item	Apply \autoref{thm:limit_allbut1}; $g(x) = \frac xx$ is the same as $g(x) = 1$ everywhere except at $x=0$. Thus $\ds\lim_{x\to0} g(x) = \lim_{x\to 0} 1 = 1.$
	\item	The function $f(x)$ is always 0, so $g\big(f(x)\big)$ is never defined as $g(x)$ is not defined at $x=0$. Therefore the limit does not exist.
	\item	The theorem requires that $\ds\lim_{x\to0} g(x)$ be equal to $g(0)$. They are not equal, so the conditions of the theorem are not satisfied, and hence the theorem is not violated.
\end{enumext}}

\exercise{Verify $\ds\lim_{x\to 0} \frac{\cos x -1}{x} =0$ \quad Hint: Multiply by $\dfrac{\cos x +1}{\cos x +1}$}{We find $\ds\lim_{x\to0}\frac{\cos^2 x-1}{x(\cos x+1)}=\lim_{x\to0}\frac{\sin^2x}{x(\cos x+1)}=\lim_{x\to0}\frac{\sin x}x\lim_{x\to0}\frac{\sin x}{\cos x+1}=0$.}

\exercise{Suppose $ax^2+bx+c=0$, so that the quadratic formula tells us the roots are given by $x=\frac{-b\pm\sqrt{b^2-4ac}}{2a}$.  Find the limiting values of the roots
\begin{enumext}[start=1]
\item when $b\to 0$
\item when $c\to 0$
\item when $a\to 0$
\end{enumext}}{\begin{enumext}[start=1]
\item $\pm\sqrt{-\frac ca}$
\item 0 and $-\frac ba$
\item $-\frac cb$ and undefined
\end{enumext}}
