\printconcepts

\exercise{Describe what an ``extreme value'' of a function is in your own words.}{Answers will vary.}

\exercise{Sketch the graph of a function $f$ on $(-1,1)$ that has both a maximum and minimum value.}{Answers will vary.}

\exercise{Describe the difference between absolute and relative maxima in your own words.}{Answers will vary.}

\exercise{Sketch the graph of a function $f$ where $f$ has a relative maximum at $x=1$ and $\fp(1)$ is undefined.}{Answers will vary.}

\exercise{T/F: If $c$ is a critical value of a function $f$, then $f$ has either a relative maximum or relative minimum at $x=c$. }{F}

\exercise{Fill in the blanks: The critical points of a function $f$ are found where $\fp(x)$ is equal to \underline{\hskip.5in} or where $\fp(x)$ is \underline{\hskip.5in}.}{Where $\fp(x)$ is equal to \underline{0} or where $\fp(x)$ is \underline{undefined}.}

\printproblems

\input{exercises/03-01-exset-01}

\input{exercises/03-01-exset-02}

\input{exercises/03-01-exset-03}

\exercise{\mbox{}\\[-2.5\baselineskip]\parbox[t]{\linewidth}{\begin{enumext}
\item Sketch the graph of a function that has a local minimum at 3 and is differentiable at 3.
\item Sketch the graph of a function that has a local minimum at 3 and is continuous but not differentiable at 3.
\item Sketch the graph of a function that has a local minimum at 3 and is not continuous at 3.
\end{enumext}}}{Answers will vary.}

\exercise{Show that 4 is a critical number of $f(x)=(x-4)^3+7$ but $f$ does not have a relative extreme value at 4.}{$\fp(x)=3(x-4)^2$, so $\fp(4)=0$.  But $f(x)>7$ for $x>4$ and $f(x)<7$ for $x<4$.}

\exercise{A cubic function is a polynomial of degree 3; that is, it has the form $ax^3+bx^2+cx+d$, where $a\neq 0$.
\begin{enumext}
\item Show that a cubic function can have 2, 1, or 0 critical numbers.  Give examples and sketches to illustrate the 3 possibilities.
\item How many local extreme values can a cubic function have?
\end{enumext}}{\mbox{}\\[-2\baselineskip]\parbox[t]{\linewidth}{\begin{enumext}\item $x^3-x$, $x^3$, and $x^3+x$ have 2, 1, and 0 critical numbers respectively.  Because the derivative is a quadratic with at most 2 roots, a cubic cannot have 3 or more critical numbers.
\item A cubic can only have 2 or 0 extreme values.
\end{enumext}}}

\exercise{Suppose that $a$ and $b$ are positive numbers.  Find the extreme values of $f(x)=x^a(1-x)^b$ on $[0,1]$.}{min: $(0,0)$ and $(1,0)$\\ max: $(\frac a{a+b},\frac{a^ab^b}{(a+b)^{a+b}})$}

\printreview

\exercise{Find $\dfrac{\dd y}{\dd x}$, where $x^2y-y^2x = 1$.}{$\frac{\dd y}{\dd x} = \frac{y (y-2 x)}{x (x-2 y)}$}

\exercise{Find the equation of the line tangent to the graph of $x^2+y^2+xy=7$ at the point $(1,2)$.}{$y=-\frac45(x-1)+2$}

\exercise{Let $f(x) = x^3+x$. 

Evaluate $\ds \lim_{s\to 0} \frac{f(x+s)-f(x)}{s}$.}{$3x^2+1$}
