\printconcepts

\exercise{T/F: Given a differentiable function $y=f(x)$, we are generally free to choose a value for $dx$, which then determines the value of $dy$.}{T}

\exercise{T/F: The symbols ``$dx$'' and ``\dx'' represent the same concept.}{T}

\exercise{T/F: The symbols ``$dy$'' and ``\dy'' represent the same concept.}{F}

\exercise{T/F: Differentials are important in the study of integration.}{T}

\exercise{How are differentials and tangent lines related?}{Answers will vary.}

\printproblems

\input{exercises/04_04_exset_01}

\input{exercises/04_04_exset_02}

\exercise{Show that the linearization of $f(x)=(1+x)^k$ at $x=0$ is $L(x)=1+kx$.}{}

\exerciseset{In Exercises}{, use the linear approximation $(1+x)^k\approx 1+kx$ to find an approximation for the function $f(x)$ for values of $x$ near zero.}{

\exercise{$f(x)=(1-x)^6$}{$1-6x$}

\exercise{$f(x)=\dfrac{2}{1-x}$}{$2+2x$}

\exercise{$f(x)=\dfrac{1}{\sqrt{1+x}}$}{$1-\frac x2$}

\exercise{$f(x)=\sqrt{2+x^2}$}{$\sqrt2$}

\exercise{$f(x)=(4+3x)^{1/3}$}{$2^{2/3}+\frac x{2^{4/3}}$}

\exercise{$\ds f(x)=\sqrt[3]{\biggl(1-\frac{1}{2+x}\biggr)^2}$}{$\frac1{2^{2/3}}+\frac x{3\cdot 2^{2/3}}$}

}

\exercise{A set of plastic spheres are to be made with a diameter of 1cm. If the manufacturing process is accurate to 1mm, what is the propagated error in volume of the spheres?}{$dV = \pm 0.157$}

\exercise{The distance, in feet, a stone drops in $t$ seconds is given by $d(t) = 16t^2$. The depth of a hole is to be approximated by dropping a rock and listening for it to hit the bottom. What is the propagated error if the time measurement is accurate to $2/10^{\text{ths}}$ of a second and the measured time is:
	\begin{enumerate}
	\item 2 seconds?
	\item	5 seconds?
	\end{enumerate}}{\mbox{}\\[-2\baselineskip]\begin{enumerate}
\item		$\pm$12.8 feet
\item		$\pm$32 feet
\end{enumerate}}

\exercise{What is the propagated error in the measurement of the cross sectional area of a circular log if the diameter is measured at $15''$, accurate to $1/4''$?
}{$\pm15\pi/8 \approx \pm5.89$in$^2$}

\exercise{A wall is to be painted that is exactly $8'$ high and is measured to be $10',\ 7''$ long. Find the propagated error in the measurement of the wall's surface area if the measurement of the length is accurate to $1/2''$. }{$\pm$48in$^2$, or 1/3ft$^2$}

\input{exercises/04_04_exset_03}

%\printreview

%\exercise{Consider $f(x) = x^2-3x+5$ on $[-1,2]$; find $c$ guaranteed by the Mean Value Theorem.}{$c=1/2$}

%\exercise{Consider $f(x) = \sin x$ on $[-\pi/2,\pi/2]$; find $c$ guaranteed by the Mean Value Theorem.}{$c=\pm \cos^{-1}(2/\pi)$}
