\printconcepts

\exercise{How are definite and indefinite integrals related?}{Answers will vary.}

\exercise{What constant of integration is most commonly used when evaluating definite integrals?}{0}

\exercise{T/F: If $f$ is a continuous function, then $\ds F(x) = \int_a^x f(t)\ dt$ is also a continuous function.}{T}

\exercise{The definite integral can be used to find ``the area under a curve.'' Give two other uses for definite integrals.}{Answers will vary.}

\printproblems

\input{exercises/05_04_exset_01}

\exercise{Explain why:
\begin{enumerate}
\item		$\ds \int_{-1}^1 x^n\ dx=0$, when $n$ is a positive, odd integer, and 
\item		$\ds \int_{-1}^1 x^n\ dx = 2\int_{0}^1 x^n\ dx$ when $n$ is a positive, even integer.
\end{enumerate}}{Explanations will vary. A sketch will help.}

\exercise{Explain why $\ds \int_{a}^{a+2\pi} \sin t\ dt = 0$ for all values of $a$.}{$\int_{a}^{a+2\pi} \sin t\ dt = \cos(a+2\pi)-\cos(a)$. Since cosine is periodic with period $2\pi$, $\cos(a+2\pi) = \cos(a)$, and hence the integral is 0.}

\input{exercises/05_04_exset_02}

\input{exercises/05_04_exset_03}

\input{exercises/05_04_exset_04}

\input{exercises/05_04_exset_05}

%\input{exercises/05_04_exset_06}

\input{exercises/05_04_exset_07}

\exercise{Let $g(x)=\ds\int_0^x f(t)\,dt$ where $f$ is the function whose graph is shown below.
\begin{enumerate}
\item Evaluate $g(x)$ for $x=0,1,2,3,4,5,6$.
\item Estimate $g(7)$.
\item Where does $g$ have a minimum value? a maximum value?
\item Sketch the graph of $g$.
\end{enumerate}

\begin{tikzpicture}
 \begin{axis}[width=.5\textwidth,tick label style={font=\scriptsize},
  axis y line=middle, axis x line=middle, name=myplot, axis on top,
  xmin=-1,xmax=7,ymin=-2,ymax=4, smooth]
  \addplot [smooth,thick,draw={\colorone},domain=0:2] {1-x};
  \addplot [smooth,thick,draw={\colorone},domain=2:6] {x-3};
  \addplot [smooth,thick,draw={\colorone},domain=6:7] {10*x-x*x-21};
  \draw [help lines,step=100] (axis cs:0,-2) grid (axis cs:7,4);
  % why steps=100?  I have no idea
 \end{axis}
 \node [right] at (myplot.right of origin) {\scriptsize $x$};
 \node [above] at (myplot.above origin) {\scriptsize $y$};
\end{tikzpicture}}{\mbox{}\\[-2\baselineskip]\begin{enumerate}
\item
\begin{tabular}{|c||c|c|c|c|c|c|c|}\hline
x & 0 & 1 & 2 & 3 & 4 & 5 & 6 \\\hline
g(x) & 0 & .5 & 0 & -.5 & 0 & 1.5 & 4 \\\hline
\end{tabular}
\item $g(7)\approx 5.7$
\item min at $x=3$; max at $x=7$
\item Approximately\\
\begin{tikzpicture}
 \begin{axis}[width=.4\textwidth,tick label style={font=\scriptsize},
  axis y line=middle, axis x line=middle, name=myplot, axis on top,
  xmin=-1,xmax=7,ymin=-1,ymax=6, smooth]
  \addplot [smooth,thick,draw={\colorone},domain=0:2] {x-x*x/2};
  \addplot [smooth,thick,draw={\colorone},domain=2:6] {x*x/2-3*x+4};
  \addplot [smooth,thick,draw={\colorone},domain=6:7] {5*x*x-x*x*x/3-21*x+22};
%  \draw [help lines,step=100] (axis cs:0,-1) grid (axis cs:7,6);
 \end{axis}
 \node [right] at (myplot.right of origin) {\scriptsize $x$};
 \node [above] at (myplot.above origin) {\scriptsize $y$};
\end{tikzpicture}
\end{enumerate}}

\exercise{For any $x>0$, define
\[g(x)=\int_1^x\frac1t\ dt.\]
\begin{enumerate}
\item Show that $g(x)$ is continuous and differentiable on $(0,\infty)$ with $g'(x)=\frac1x$.
\item Show that for any positive $x$ and $y$ we have $g(xy)=g(x)+g(y)$.\\{}
[Hint: Treat $y$ as a constant, consider the derivative with respect to $x$ of each side of the proposed equation, and apply \autoref{thm:antideriv_const}.]
\item Show that for any positive $x$ and any $r$ we have $g(x^r)=rg(x)$.\\{}
[Hint: Consider the derivative with respect to $x$ of each side of the proposed equation and apply \autoref{thm:antideriv_const}.]
\end{enumerate}}{\mbox{}\\[-2\baselineskip]\begin{enumerate}
\item This is a consequence of \autoref{thm:FTC1}.
\item The derivative of the left is $g'(xy)y=\frac1{xy}y$.  The derivative of the right is $g'(x)=\frac1x$. \autoref{thm:antideriv_const} tells us that the left and right therefore differ by a constant.  Looking at $x=1$ and $g(1)=0$ tells us that this difference is 0.
\item The derivative of the left is $g'(x^r)rx^{r-1}=\frac1{x^r}rx^{r-1}$.  The derivative of the right is $rg'(x)=r\frac1x$. \autoref{thm:antideriv_const} tells us that the left and right therefore differ by a constant.  Looking at $x=1$ and $g(1)=0$ tells us that this difference is 0.
\end{enumerate}}

%\printreview

%\input{exercises/05_04_exset_07}
