\printconcepts

\exercise{What is ``total signed area''?}{Answers will vary.}

\exercise{What is ``displacement''?}{Answers will vary.}

\exercise{What is $\ds \int_3^3 \sin x\ dx$?}{0}

\exercise{Give a single definite integral that has the same value as 

$\ds \int_0^1(2x+3)\ dx + \int_1^2 (2x+3)\ dx$.}{$\int 0^2 (2x+3)\ dx$}

\printproblems

\input{exercises/05_02_exset_01}

\input{exercises/05_02_exset_02}

\input{exercises/05_02_exset_03}

\exercise{An object is thrown straight up with a velocity, in ft/s, given by $v(t) = -32t+64$, where $t$ is in seconds, from a height of 48 feet.
\begin{enumerate}
\item		What is the object's maximum velocity?
\item		What is the object's maximum displacement?
\item		When does the maximum displacement occur?
\item		When will the object reach a height of 0? (Hint: find when the displacement is $-48$ft.)
\end{enumerate}}{\begin{enumerate}
\item		64ft/s
\item		64ft
\item		$t=2$
\item		$t=2+\sqrt{7}\approx 4.65$ seconds
\end{enumerate}}

\exercise{An object is thrown straight up with a velocity, in ft/s, given by $v(t) = -32t+96$, where $t$ is in seconds, from a height of 64 feet.
\begin{enumerate}
\item		What is the object's initial velocity?
\item		When is the object's displacement 0?
\item		How long does it take for the object to return to its initial height?
\item		When will the object reach a height of 210 feet?
\end{enumerate}}{\begin{enumerate}
\item		96ft/s
\item		6 seconds
\item		6 seconds
\item		Never; the maximum height is 208ft.
\end{enumerate}}

\input{exercises/05_02_exset_04}

\input{exercises/05_02_exset_05}

\printreview

\input{exercises/05_02_exset_06}
