\printconcepts

\exercise{A fundamental calculus technique is to use \underline{\hskip .5in} to refine approximations to get an exact answer.}{limits}

\exercise{What is the upper bound in the summation $\ds \sum_{i=7}^{14} (48i-201)$?}{14}

\exercise{This section approximates definite integrals using what geometric shape?}{Rectangles.}

\exercise{T/F: A sum using the Right Hand Rule is an example of a Riemann Sum.}{T}

\printproblems

\input{exercises/05_03_exset_01}

\input{exercises/05_03_exset_02}

\input{exercises/05_03_exset_03}

\input{exercises/05_03_exset_04}

\ifthenelse{\boolean{printquestions}}{\columnbreak}{}

\input{exercises/05_03_exset_05}

\input{exercises/05_03_exset_06}

\ifthenelse{\boolean{printquestions}}{\columnbreak}{}

\exercise{Use six rectangles to approximate the area under the given graph of $f$ from $x=0$ to $x=12$, using:
\begin{enumerate}
\item The Left Hand Rule,
\item The Right Hand Rule,
\item The Midpoint Rule.
\end{enumerate}
\begin{tikzpicture}
\begin{axis} [width=.4\textwidth,%\marginparwidth+25pt,
tick label style={font=\scriptsize},axis y line=middle,axis x line=middle,
name=myplot,axis on top,%all axes={grid},
                        ymin=0,ymax=10,
                        xmin=0,xmax=12,smooth]
   \addplot [{\coloronefill},fill={\coloronefill},area style,domain=0:12] {9-(x/4)^2} \closedcycle;
   \addplot [smooth,thick,{\colorone},domain=0:12] {9-(x/4)^2};
   \draw [help lines,step=10] (axis cs:0,0) grid (axis cs:12,10);
   % steps=10 was guess and check
\end{axis}
\node [right] at (myplot.right of origin) {\scriptsize $x$};
\node [above] at (myplot.above origin) {\scriptsize $y$};
\end{tikzpicture}}{\begin{enumerate}
\item	Exact expressions will vary; $80.5$.
\item	$72.25$
\item	$62.5$
\end{enumerate}}

\exercise{A car accelerates from 0 to 40 mph in 30 seconds. The speedometer reading at each 5 second interval during this time is given in the table below. Estimate how far the car travels during this 30 second period using the velocities at:
\begin{enumerate}
\item The beginning of each time interval.
\item The end of each time interval.
\end{enumerate}

\begin{center}
\begin{tabular} {|c|c|c|c|c|c|c|c|}
\hline
$\mathbf{t}$ (sec)&0&5&10&15&20&25&30\\ \hline
$\mathbf{v}$ (mph)&0&6&14&23&30&36&40\\ \hline
\end{tabular}
\end{center}}{\begin{enumerate}
\item	$(5\text{ s})((0+6+14+23+30+36)\text{ mph})=545\frac{\text{mi s}}{\text{hr}}\times\frac{1\text{ hr}}{3600\text{ s}}\times{5280\text{ ft}}{1\text{ mi}}=799\text{ ft}$
\item	$(5\text{ s})((6+14+23+30+36+40)\text{ mph})=585\frac{\text{mi s}}{\text{hr}}\times\frac{1\text{ hr}}{3600\text{ s}}\times{5280\text{ ft}}{1\text{ mi}}=858\text{ ft}$
\end{enumerate}}

\exercise{Use Theorems \ref{thm:summation} and \ref{thm:riemannSum} to justify the remaining property in Theorem \ref{thm:defintprop}: \[\int_a^b k\cdot f(x)\,dx =k\int_a^b f(x)\,dx\]}{\begin{align*}
\int_a^b k\cdot f(x)\,dx
&=\lim_{n\to\infty}\sum_{i=1}^n k\cdot f(c_i)\Delta x
\quad\text{T\ref{thm:riemannSum}.2} \\
&=\lim_{n\to\infty}k\cdot\sum_{i=1}^n k\cdot f(c_i)\Delta x
\quad\text{T\ref{thm:summation}.3} \\
&=k\cdot\lim_{n\to\infty}\sum_{i=1}^n k\cdot f(c_i)\Delta x
\quad\text{T\ref{thm:limit_algebra}.4} \\
&=k\int_a^b f(x)\,dx\quad\text{T\ref{thm:riemannSum}.2}
\end{align*}}

\exercise{Use Theorems \ref{thm:summation} and \ref{thm:riemannSum} to justify the remaining property in Theorem \ref{thm:further_def_int_props}: If $f(x)\leq M$ for all $x$ in $[a,b]$, then \[\int_a^b f(x)\,dx\leq M(b-a).\]}{Let $f$ and $M$ be as given.
\begin{align*}
\int_a^b f(x)\,dx
&=\lim_{n\to\infty}\sum_{i=1}^n f(c_i)\Delta x
\quad\text{T\ref{thm:riemannSum}.2} \\
&\le\lim_{n\to\infty}\sum_{i=1}^n M\Delta x \\
&=\int_a^b M\,dx\quad\text{T\ref{thm:riemannSum}.2} \\
&=M(b-a)
\end{align*}}

\printreview

\input{exercises/05_03_exset_07}
