\printconcepts

\exercise{The definite integral was defined with what two stipulations?}{The interval of integration is finite, and the integrand is continuous on that interval.}

\exercise{If $\ds \lim_{b\to \infty} \int_0^b f(x)\dd x$ exists, then the integral $\ds \int_0^\infty f(x)\dd x$ is said to \underline{\hskip 1in}.}{converge}

\exercise{If $\ds \int_1^\infty f(x)\dd x=10$, and $0\leq g(x)\leq f(x)$ for all $x$, then we know that $\ds \int_1^\infty g(x)\dd x$  \underline{\hskip 1in}.}{converges; could also state $\le10$.}

\exercise{For what values of $p$ will $\ds \int_1^\infty \frac1{x^p}\dd x$ converge?}{$p>1$}

\exercise{For what values of $p$ will $\ds \int_{10}^\infty \frac1{x^p}\dd x$ converge?}{$p>1$}

\exercise{For what values of $p$ will $\ds \int_{0}^1 \frac1{x^p}\dd x$ converge?}{$p<1$}

\printproblems

\input{exercises/06-07-exset-01}

\input{exercises/06-07-exset-02}

%\input{exercises/06-05-exset-03}
%\input{exercises/06-05-exset-04}
%\input{exercises/06-05-exset-05}
%\printreview
%\input{exercises/05-04-exset-07}

\exercise{In probability theory, the lifetimes of certain devices (e.g. certain types of fuses and light bulbs) are modeled by an \emph{Exponential Distribution}.
\begin{enumext}[start=1]
\item The probability that a device lasts more than $a$ (time units) is $\int_a^\infty\lambda e^{-\lambda x}\dd x$ where $\lambda$ is a parameter that depends on the type of device. Evaluate this integral.
\item The expected lifetime of the device is given by $\int_0^\infty x\lambda e^{-\lambda x}\dd x$. Evaluate this integral.
\item What is the probability that a given device lasts more than the expected lifetime for such devices?
\end{enumext}}{\mbox{}\\[-2\baselineskip]\parbox[t]{\linewidth}{\begin{enumext}[start=1]
\item $e^{-\lambda a}$
\item $\frac1\lambda$
\item $e^{-1}$
\end{enumext}}}

\exercise{For $n>0$, the gamma function\index{gamma function} is defined by $\ds\Gamma(n)=\int_0^\infty x^{n-1}e^{-x}\dd x$.
\begin{enumext}[start=1]
\item Show that $\Gamma(1)=1$.
\item Show that $\Gamma(n+1)=n\Gamma(n)$ for $n>1$.
\item Conclude that $\Gamma(n+1)=n!$ for integers $n$ such that $n\ge 1$.
\item Show that this converges for $0<n<1$.
\end{enumext}}{}
