\printconcepts

\exercise{Axes drawn in space must conform to the \underline{\hskip .5in} \underline{\hskip .5in} rule.}{right hand}

\exercise{In the plane, the equation $x=2$ defines a \underline{\hskip.5in}; in space, $x=2$ defines a \underline{\hskip .5in}.}{line; plane}

\exercise{In the plane, the equation $y=x^2$ defines a \underline{\hskip.5in}; in space, $y=x^2$ defines a \underline{\hskip .5in}.}{curve (a parabola); surface (a cylinder)}

\exercise{Which quadric surface looks like a Pringles\textregistered\ chip?}{a hyperbolic paraboloid}

\exercise{Consider the hyperbola $x^2-y^2=1$ in the plane. If this hyperbola is rotated about the $x$-axis, what quadric surface is formed?}{a hyperboloid of two sheets}

\exercise{Consider the hyperbola $x^2-y^2=1$ in the plane. If this hyperbola is rotated about the $y$-axis, what quadric surface is formed?}{a hyperboloid of one sheet}

\printproblems

\exercise{The points $A=(1,4,2)$, $B=(2,6,3)$ and $C=(4,3,1)$ form a triangle in space. Find the distances between each pair of points and determine if the triangle is a right triangle.}{$\norm{\overline{AB}} = \sqrt{6}$; $\norm{\overline{BC}} = \sqrt{17}$; $\norm{\overline{AC}} = \sqrt{11}$. Yes, it is a right triangle as $\norm{\overline{AB}}^2+\norm{\overline{AC}}^2=\norm{\overline{BC}}^2$.}

\exercise{The points $A=(3,2,0)$, $B=(1,0,1)$, and $C=(-1,1,3)$ form a triangle in space.  Find the distance between each pair of points and determine if the triangle is isosceles.}{$\norm{\overline{AB}}=3$, $\norm{\overline{AC}}=\sqrt{26}$, $\norm{\overline{BC}}=3$. The triangle is isosceles.}

\exercise{Explain why three points lie on a line if and only if the distance between two of the points is equal to the sum of the distances from each of these points to the third.}{}

\exercise{Determine whether or not the points $A=(2,0,3)$, $B=(0,-3,4)$, and $C=(4,3,2)$ lie on a line.}{$\norm{\overline{AB}}=\sqrt{14}$, $\norm{\overline{AC}}=\sqrt{14}$, $\norm{\overline{BC}}=2\sqrt{14}$.  The points lie on a line.}

\exercise{Determine whether or not the points $A=(4,5,3)$, $B=(6,8,7)$, and $C=(0,0,-5)$ lie on a line.}{$\norm{\overline{AB}}=\sqrt{29}$, $\norm{\overline{AC}}=\sqrt{105}$, $\norm{\overline{BC}}=2\sqrt{61}$.  The points do not lie on a line.}

\exercise{The points $A=(1,1,3)$, $B=(3,2,7)$, $C=(2,0,8)$ and $D = (0,-1,4)$ form a quadrilateral $ABCD$ in space. Is this a parallelogram?}{Yes, as opposite sides have equal length. $\norm{\overline{AB}} = \sqrt{21}=\norm{\overline{CD}}$; $\norm{\overline{BC}} = \sqrt{6}=\norm{\overline{AD}}$.}

\exercise{Find the center and radius of the sphere defined by\\
$x^2-8x+y^2+2y+z^2+8=0.$}{Center at $(4,-1,0)$; radius = 3}

\exercise{Find the center and radius of the sphere defined by\\
$x^2+y^2+z^2+4x-2y-4z+4=0.$}{Center at $(-2,1,2)$; radius = $\sqrt{5}$}

\exercise{Show that the point $A=(2,1,5)$ is inside the sphere given by
\[x^2+y^2+z^2-2x+4y-6z=11.\]
Is $A$ closer to the center of the sphere or to the surface of the sphere?}{closer to the surface}
% (1,-2,3) radius=5
% d( center, A ) = sqrt(14) ~ 3.7

\exercise{Let $P = (x_1, y_1, z_1)$ and $Q = (x_2, y_2, z_2)$ be points in space.  Show that the midpoint of $\overline{PQ}$ is 
\[\left(\frac{x_1+x_2}2,\frac{y_1+y_2}2,\frac{z_1+z_2}2\right).\]}{}

\input{exercises/10_01_exset_01}

\input{exercises/10_01_exset_02}

\input{exercises/10_01_exset_05}

\input{exercises/10_01_exset_03}

\input{exercises/10_01_exset_04}
