\printconcepts

\exercise{The dot product of two vectors is a \underline{\hskip .5in}, not a vector.}{Scalar}

\exercise{How are the concepts of the dot product and vector magnitude related?}{The magnitude of a vectors is the square root of the dot product of a vector with itself; that is, $\norm{\vec v} = \sqrt{\vec v\cdot \vec v}$.}

\exercise{How can one quickly tell if the angle between two vectors is acute or obtuse?}{By considering the sign of the dot product of the two vectors. If the dot product is positive, the angle is acute; if the dot product is negative, the angle is obtuse.}

\exercise{Give a synonym for ``orthogonal.''}{``Perpendicular'' is one answer.}

\printproblems

\input{exercises/10_03_exset_01}

\exercise{Create your own vectors $\vec u$, $\vec v$ and $\vec w$ in $\mathbb{R}^2$ and show that $\vec u\cdot (\vec v+\vec w) = \vec u\cdot \vec v + \vec u\cdot \vec w$.}{Answers will vary.}

\exercise{Create your own vectors $\vec u$ and $\vec v$  in $\mathbb{R}^3$  and scalar $c$ and show that $c(\vec u\cdot \vec v) = \vec u\cdot (c\vec v)$.}{Answers will vary.}

\input{exercises/10_03_exset_02}

\input{exercises/10_03_exset_03}

\input{exercises/10_03_exset_04}

\input{exercises/10_03_exset_05}

\exercise{A 10lb box sits on a ramp that rises 4ft over a distance of 20ft. How much force is required to keep the box from sliding down the ramp?}{1.96lb}

\exercise{A 10lb box sits on a 15ft ramp that makes a $30^\circ$ angle with the horizontal. How much force is required to keep the box from sliding down the ramp?}{5lb}

\exercise{How much work is performed in moving a box horizontally 10ft with a force of 20lb applied at an angle of $45^\circ$ to the horizontal?}{$141.42$ft--lb}

\exercise{How much work is performed in moving a box horizontally 10ft with a force of 20lb applied at an angle of $10^\circ$ to the horizontal?}{$196.96$ft--lb}

\exercise{How much work is performed in moving a box up the length of a ramp that rises 2ft over a distance of 10ft, with a force of 50lb applied horizontally?}{$500$ft--lb}

\exercise{How much work is performed in moving a box up the length of a ramp that rises 2ft over a distance of 10ft, with a force of 50lb applied at an angle of $45^\circ$ to the horizontal?}{$424.26$ft--lb}

\exercise{How much work is performed in moving a box up the length of a 10ft ramp that makes a $5^\circ$ angle with the horizontal, with 50lb of force applied in the direction of the ramp?}{$500$ft--lb}

\exercise{For any two vectors $\vec{u}$ and $\vec{v}$ use the properties of the dot product to show that
\[
\norm{\vec{u}\pm\vec{v}}^2=\norm{\vec{u}}^2+\norm{\vec{v}}^2\pm2(\vec{u}\cdot\vec{v}).
\]}{}

\exercise{For any two vectors $\vec{u}$ and $\vec{v}$ show that
\[
\norm{\vec{u}+\vec{v}}^2+\norm{\vec{u}-\vec{v}}^2=2\norm{\vec{u}}^2+2\norm{\vec{v}}^2.
\]
Interpret this as a statement about parallelograms.}{}

\exercise{Consider two nonzero vectors $\vec{u}$ and $\vec{v}$ and the angle between them $\theta$.  The vectors $\vec{u}$, $\vec{v}$, and $\vec{u} - \vec{v}$ form the triangle as follows.
\begin{center}
\begin{tikzpicture}[scale=2]
    \draw[thick,->] (0,0) node[anchor=south west]{$\ \ \ \theta$} -- node[anchor=north]{$\vec{u}$} (2,0);
    \draw[thick,->] (0,0) -- node[anchor=south]{$\vec{v}$} (1,1);
    \draw[thick,->, dashed] (1,1) -- node[anchor=south west]{$\vec{u} - \vec{v}$} (2,0);
    \draw[->,thin] (3/5,0) arc [start angle=0, end angle=45, radius=3/5];
\end{tikzpicture}
\end{center}
\begin{enumerate}
\item Use the Law of Cosines to show that
\[
\norm{\vec{u}-\vec{v}}^2
=\norm{\vec{u}}^2+\norm{\vec{v}}^2-2\norm{\vec{u}}\norm{\vec{v}}\cos\theta.
\]

\item  Use (a) and the previous problem to conclude the formula 
\[\vec{u}\cdot\vec{v}=\norm{\vec{u}}\norm{\vec{v}}\cos\theta.\]
\end{enumerate}}{}

\exercise{Suppose we know that $\norm{\vec{u}}=5$, $\norm{\vec{v}}=4$, and the angle between $\vec{u}$ and $\vec{v}$ is $\theta=\pi/3$. Determine the following.
\begin{enumerate}
\item $\norm{\vec{u}+\vec{v}}$.
\item $\vec{u}\cdot\vec{v}$.
\item $\norm{\frac12\vec{u}+3\vec{v}}$.
\item $\norm{\vec{u}-\vec{v}}$.
\end{enumerate}}{}

\exercise{Show that the two diagonals of a parallelogram intersect in right angles if and only if all four sides of the parallelogram have the same length.}{}

\exercise{Show that for any two vectors $\vec{u}$ and $\vec{v}$ we have
\[\abs{\vec{u}\cdot\vec{v}}\leq\norm{\vec{u}}\norm{\vec{v}}.\]
This is called the \emph{Cauchy-Schwarz inequality}.}{}

\exercise{Show that for any two vectors $\vec{u}$ and $\vec{v}$ we have
\[\norm{\vec{u}+\vec{v}}\leq\norm{\vec{u}}+\norm{\vec{v}}.\]
This is called the \emph{triangle inequality}.  Explain the name.}{}
