\printconcepts

\exercise{Name two different things that cannot be described with just one number, but rather need 2 or more numbers to fully describe them.}{Answers will vary.}

\exercise{What is the difference between $(1,2)$ and $\bracket{1,2}$?}{$(1,2)$ is a point; $\bracket{1,2}$ is a vector that describes a displacement of 1 unit in the $x$-direction and 2 units in the $y$-direction.}

\exercise{What is a unit vector?}{A vector with magnitude 1.}

\exercise{What does it mean for two vectors to be parallel?}{Their respective unit vectors are parallel; unit vectors $\vec u_1$ and $\vec u_2$ are parallel if $\vec u_1 = \pm \vec u_2$.}

\exercise{What effect does multiplying a vector by $-2$ have?}{It stretches the vector by a factor of 2, and points it in the opposite direction.}

\printproblems

\input{exercises/10_02_exset_01}

\exercise{Let $\vec u =\bracket{1,-2}$ and $\vec v=\bracket{1,1}$. 
\begin{enumerate}
	\item Find $\vec u+\vec v$, $\vec u-\vec v$, $2\vec u-3\vec v$.
	\item	Sketch the above vectors on the same axes, along with $\vec u$ and $\vec v$.
	\item		Find $\vec x$ where $\vec u+\vec x = 2\vec v-\vec x$.
\end{enumerate}
}{\begin{enumerate}
	\item $\vec u+\vec v =\bracket{2,-1}$; $\vec u -\vec v =\bracket{0,-3}$; $2\vec u-3\vec v =\bracket{-1,-7}$.
	\item	[(c)]	$\vec x =\bracket{1/2,2}$.
\end{enumerate}}

\exercise{Let $\vec u =\bracket{1,1,-1}$ and $\vec v=\bracket{2,1,2}$. 
\begin{enumerate}
	\item Find $\vec u+\vec v$, $\vec u-\vec v$, $\pi\vec u-\sqrt{2}\vec v$.
	\item	Sketch the above vectors on the same axes, along with $\vec u$ and $\vec v$.
	\item		Find $\vec x$ where $\vec u+\vec x = \vec v+2\vec x$.
\end{enumerate}
}{\begin{enumerate}
	\item $\vec u+\vec v =\bracket{3,2,1}$; $\vec u -\vec v =\bracket{-1,0,-3}$; $\pi\vec u-\sqrt{2}\vec v =\bracket{\pi-2\sqrt{2},\pi-\sqrt{2},-\pi-2\sqrt{2}}$.
	\item	[(c)]	$\vec x =\bracket{-1,0,-3}$.
\end{enumerate}}

\input{exercises/10_02_exset_02}

\input{exercises/10_02_exset_03}

\exercise{Under what conditions is $\norm{\vec u}+\norm{\vec v} = \norm{\vec u+\vec v}$?}{When $\vec u$ and $\vec v$ have the same direction. (Note: parallel is not enough.)}

\input{exercises/10_02_exset_04}

\exercise{Find the unit vector in the first quadrant of $\mathbb{R}^2$ that makes a $50^\circ$ angle with the $x$-axis.}{$\vec u =\bracket{\cos 50^\circ,\sin 50^\circ}\approx\bracket{0.643,0.766}$.}

\exercise{Find the unit vector in the second quadrant of $\mathbb{R}^2$ that makes a $30^\circ$ angle with the $y$-axis.}{$\vec u =\bracket{\cos 120^\circ,\sin 120^\circ}=\bracket{-1/2,\sqrt{3}/2}$.}

\exercise{Verify, from \autoref{idea:unit_vectors}, that $\vec u=\bracket{\cos\theta\sin\phi,\sin\theta\sin\phi,\cos\phi}$ is a unit vector for all angles $\theta$ and $\phi$.}{\mbox{}\\[-2\baselineskip]\begin{align*}
	\norm{\vec u}
	&=\sqrt{\cos^2\theta\sin^2\phi+\sin^2\theta\sin^2\phi+\cos^2\phi}\\
	&=\sqrt{\sin^2\phi(\cos^2\theta+\sin^2\theta)+\cos^2\phi}\\
	&=\sqrt{\sin^2\phi+\cos^2\phi}\\
	&=1.
\end{align*}}

\input{exercises/10_02_exset_05}

\input{exercises/10_02_exset_06}

\exercise{Let $P = (x_1, y_1, z_1)$ and $Q = (x_2, y_2, z_2)$ be points in space.  Let $M$ be the midpoint of $\overline{PQ}$. Explain why $\overrightarrow{PM} = \overrightarrow{MQ}$ and $\overrightarrow{PM} + \overrightarrow{MQ} = \overrightarrow{{PQ}}$.  Use these facts to find the coordinates of the point $M$ (again).}{}

\exercise{Let $P=(1,2,3)$, $Q=(2,-1,4)$, and $R=(-1,6,1)$ be the three corners of a parallelogram in space.  Find the possible locations of the fourth corner.}{} % todo the solution

\exercise{Use vectors to show that the diagonals of a parallelogram bisect each other.}{}

\exercise{A \emph{median} of a triangle is a line segment from a vertex to the midpoint of the opposite side.  Show that the three medians of a triangle intersect in a single point.  This point is called the \emph{centroid} of the triangle.  Show that the distance from a vertex to the centroid is two-thirds the length of the median from this vertex.}{}
