\begin{exerciseset}{Exercises}{ ask you to solve a variety of problems based on the principles of projectile motion.}

\exercise{A boy whirls a ball, attached to a  3ft string, above his head in a counter-clockwise circle. The ball makes 2 revolutions per second.

At what $t$-values should the boy release the string so that the ball heads directly for a tree standing 10ft in front of him?
}{At $t$-values of $\sin^{-1}(9/30)/(4\pi) + n/2 \approx 0.024 + n/2$ seconds, where $n$ is an integer.}

\exercise{David faces Goliath with only a stone in a 3ft sling,  which he whirls above his head at 4 revolutions per second. They stand 20ft apart.
\begin{enumerate}
	\item At what $t$-values must David release the stone in his sling in order to hit Goliath?
	\item What is the speed at which the stone is traveling when released?
	\item Assume David releases the stone from a height of 6ft and Goliath's forehead is 9ft above the ground. What angle of elevation must David apply to the stone to hit Goliath's head?
\end{enumerate}}{The stone, while whirling, can be modeled by $\vrt =\bracket{3\cos(8\pi t),3\sin (8\pi t)}$. 
\begin{enumerate}
	\item For $t$-values $t=\sin^{-1}(3/20)/(8\pi) + n/4 \approx 0.006 + n/4$, where $n$ is an integer.
	\item $\norm{\vrp(t)} = 24\pi\approx 51.4$ft/s
	\item At $t=0.006$, the stone is approximately $19.77$ft from Goliath. Using the formula for projectile motion, we want the angle of elevation that lets a projectile starting at $\bracket{0,6}$ with a initial velocity of $51.4$ft/s arrive at $\bracket{19.77,9}$. The desired angle is $0.27$ radians, or $15.69^\circ$.
\end{enumerate}}

\exercise{A hunter aims at a deer which is 40 yards away. Her crossbow is at a height of 5ft, and she aims for a spot on the deer 4ft above the ground. The crossbow fires her arrows at 300ft/s.
\begin{enumerate}
	\item At what angle of elevation should she hold the crossbow to hit her target?
	\item	If the deer is moving perpendicularly to her line of sight at a rate of 20mph, by approximately how much should she lead the deer in order to hit it in the desired location?
\end{enumerate}}{\mbox{}\\[-2\baselineskip]\parbox[t]{\linewidth}{\begin{enumerate}
	\item Holding the crossbow at an angle of $0.013$ radians, $\approx 0.745^\circ$ will hit the target $0.4$s later. (Another solution exists, with an angle of $89^\circ$, landing $18.75$s later, but this is impractical.)
	\item	In the .4 seconds the arrow travels, a deer, traveling at 20mph or 29.33ft/s, can travel 11.7ft. So she needs to lead the deer by 11.7ft.
\end{enumerate}}}

\exercise{A baseball player hits a ball at 100mph, with an initial height of 3ft and an angle of elevation of $20^\circ$, at Boston's Fenway Park. The ball flies towards the famed ``Green Monster,'' a wall 37ft high located 310ft from home plate. 
\begin{enumerate}
	\item Show that as hit, the ball hits the wall.
	\item Show that if the angle of elevation is $21^\circ$, the ball clears the Green Monster.
\end{enumerate}}{The position function of the ball is\\
$\vrt =\bracket{(146.67\cos\theta)t,-16t^2+(146.67\sin\theta)t+3}$,\\
where $\theta$ is the angle of elevation.
\begin{enumerate}
	\item With $\theta=20^\circ$, the ball reaches 310ft from home plate in 2.25 seconds; at this time, the height of the ball is 34.9ft, not enough to clear the Green Monster.
	\item	With $\theta=21^\circ$, the ball reaches 310ft from home plate in 2.26s, with a height of 40ft, clearing the wall.
\end{enumerate}}

\exercise{A Cessna flies at 1000ft at 150mph and drops a box of supplies to the professor (and his wife) on an island. Ignoring wind resistance, how far horizontally will the supplies travel before they land?}{The position function is $\vrt =\bracket{220t,-16t^2+1000}$. The $y$-component is 0 when $t=7.9$; $\vec r(7.9) =\bracket{1739.25,0}$, meaning the box will travel about 1740ft horizontally before it lands.}

\exercise{A football quarterback throws a pass from a height of 6ft, intending to hit his receiver 20yds away at a height of 5ft.
\begin{enumerate}
	\item If the ball is thrown at a rate of 50mph, what angle of elevation is needed to hit his intended target?
	\item If the ball is thrown at with an angle of elevation of $8^\circ$, what initial ball speed is needed to hit his target?
\end{enumerate}}{The position function of the ball is\\
$\vrt =\bracket{(v_0\cos\theta)t,-16t^2+(v_0\sin\theta)t+6}$,\\
where $\theta$ is the angle of elevation and $v_0$ is the initial ball speed.
\begin{enumerate}
	\item With $v_0 = 73.33$ft/s, there are two angles of elevation possible. An angle of $\theta = 9.47^\circ$ delivers the ball in 0.83s, while an angle of $79.57^\circ$ delivers the ball in $4.5$s.
	\item	With $\theta=8^\circ$, the initial speed must be 53.8mph$\approx 78.9$ft/s.	
\end{enumerate}}

\exercisesetend
