\printconcepts

\exercise{How is \textit{velocity} different from \textit{speed}?}{Velocity is a vector, indicating an objects direction of travel and its rate of distance change (i.e., its speed). Speed is a scalar.}

\exercise{What is the difference between \textit{displacement} and \textit{distance traveled}?}{Displacement is a vector, indicating the difference between the starting and ending positions of an object. Distance traveled is a scalar, indicating the arc length of the path followed.}

\exercise{What is the difference between \textit{average velocity} and \textit{average speed}?}{The average velocity is found by dividing the displacement by the time traveled --- it is a vector. The average speed is found by dividing the distance traveled by the time traveled --- it is a scalar. }

\exercise{\textit{Distance traveled} is the same as \underline{\hskip .5in} \underline{\hskip .5in}, just viewed in a different context.}{arc length}

\exercise{Describe a scenario where an object's average speed is a large number, but the magnitude of the average velocity is not a large number.}{One example is traveling at a constant speed $s$ in a circle, ending at the starting position. Since the displacement is $\vec 0$, the average velocity is $\vec 0$, hence $\vnorm 0=0$. But traveling at constant speed $s$ means the average speed is also $s >0$.}

\exercise{Explain why it is not possible to have an average velocity with a large magnitude but a small average speed.}{Distance traveled is always greater than or equal to the magnitude of displacement, therefore average speed will always be at least as large as the magnitude of the average velocity.}

\printproblems

\input{exercises/11_03_exset_01}

\input{exercises/11_03_exset_02}

\input{exercises/11_03_exset_03}

\input{exercises/11_03_exset_04}

\input{exercises/11_03_exset_05}

\input{exercises/11_03_exset_06}

\input{exercises/11_03_exset_07}
