\printconcepts

\exercise{Explain how the vector $\vec v=\bracket{1,0,3}$ can be thought of as having a ``slope'' of 3.}{Answers will vary. The displacement of the vector is one unit in the $x$-direction and 3 units in the $z$-direction, with no change in $y$. Thus along a line parallel to $\vec v$, the change in $z$ is 3 times the change in $x$ --- i.e., a ``slope'' of 3. Specifically, the line in the $x$-$z$ plane parallel to $z$ has a slope of 3.}

\exercise{Explain how the vector $\vec v=\bracket{0.6,0.8, -2}$ can be thought of as having a ``slope'' of $-2$.}{Answers will vary. Let $\vec u =\bracket{0.6,0.8}$; this is a unit vector. The displacement of the vector is one unit in the $\vec u$-direction and $-2$ units in the $z$-direction. In the plane containing the $z$-axis and the vector $\vec u$, the line parallel to $\vec v$ has slope $-2$.}

\exercise{T/F: Let $z=f(x,y)$ be differentiable at $P$. If $\vec n$ is a normal vector to the tangent plane of $f$ at $P$, then $\vec n$ is orthogonal to $\ell_x$ and $\ell_y$ at $P$.}{T}

\exercise{Explain in your own words why we do not refer to \textit{the} tangent line to a surface at a point, but rather to \textit{directional} tangent line\textit{s} to a surface at a point.}{On a surface through a point, there are many different smooth curves, each with a tangent line at the point. Each of these tangent lines is also ``tangent'' to the surface. There is not just one tangent line, but many, each in a different direction. Therefore we refer to directional tangent lines, not just \textit{the} tangent line.}

\printproblems

\input{exercises/12-06-exset-01}

\input{exercises/12-06-exset-02}

\input{exercises/12-06-exset-03}

\input{exercises/12-06-exset-04}

\input{exercises/12-06-exset-05}

% todo include some exercises finding the distance from a surface to a point as in Example 13.7.4
