\printconcepts

\exercise{Give two examples (other than those given in the text) of ``real world'' functions that require more than one input.}{Answers will vary.}

\exercise{The graph of a function of two variables is a \underline{\hskip .5in}.}{surface}

\exercise{Most people are familiar with the concept of level curves in the context of \underline{\hskip .5in} maps.}{topographical}

\exercise{T/F: Along a level curve, the output of a function does not change.}{T}

\exercise{The analogue of a level curve for functions of three variables is a level \underline{\hskip .5in}.}{surface}

\exercise{What does it mean when level curves are close together? Far apart?}{When level curves are close together, it means the function is changing $z$-values rapidly. When far apart, it changes $z$-values slowly.}

\printproblems

\input{exercises/12_01_exset_01}

\input{exercises/12_01_exset_02}

\input{exercises/12_01_exset_03}

\input{exercises/12_01_exset_04}

\exercise{Compare the level curves of Exercises \ref{12_01_ex_21} and \ref{12_01_ex_22}. How are they similar, and how are they different? Each surface is a quadric surface; describe how the level curves are consistent with what we know about each surface.}{The level curves for each surface are similar; for $z=\sqrt{x^2+4y^2}$ the level curves are ellipses of the form $\frac{x^2}{c^2}+\frac{y^2}{c^2/4}=1$, i.e., $a=c$ and $b=c/2$; whereas for $z=x^2+4y^2$ the level curves are ellipses of the form $\frac{x^2}{c}+\frac{y^2}{c/4}=1$, i.e., $a=\sqrt{c}$ and $b=\sqrt{c}/2$. The first set of ellipses are spaced evenly apart, meaning the function grows at a constant rate; the second set of ellipses are more closely spaced together as $c$ grows, meaning the function grows faster and faster as $c$ increases.

The function $z=\sqrt{x^2+4y^2}$ can be rewritten as $z^2=x^2+4y^2$, an elliptic cone; the function $z=x^2+4y^2$ is a paraboloid, each matching the description above.}
