\printconcepts
\exercise{Explain how the vector $\vec v=\la 1,0,3\ra$ can be thought of as having a ``slope'' of 3.
}{
Answers will vary. The displacement of the vector is one unit in the $x$-direction and 3 units in the $z$-direction, with no change in $y$. Thus along a line parallel to $\vec v$, the change in $z$ is 3 times the change in $x$ -- i.e., a ``slope'' of 3. Specifically, the line in the $x$-$z$ plane parallel to $z$ has a slope of 3.
}
\exercise{Explain how the vector $\vec v=\la 0.6,0.8, -2\ra$ can be thought of as having a ``slope'' of $-2$.
}{
Answers will vary. Let $\vec u = \la 0.6,0.8\ra$; this is a unit vector. The displacement of the vector is one unit in the $\vec u$-direction and $-2$ units in the $z$-direction. In the plane containing the $z$-axis and the vector $\vec u$, the line parallel to $\vec v$ has slope $-2$.
}
\exercise{T/F: Let $z=f(x,y)$ be differentiable at $P$. If $\vec n$ is a normal vector to the tangent plane of $f$ at $P$, then $\vec n$ is orthogonal to $f_x$ and $f_y$ at $P$.
}{T
}
\exercise{Explain in your own words why we do not refer to \textit{the} tangent line to a surface at a point, but rather to \textit{directional} tangent line\textit{s} to a surface at a point.
}{On a surface through a point, there are many different smooth curves, each with a tangent line at the point. Each of these tangent lines is also ``tangent'' to the surface. There is not just one tangent line, but many, each in a different direction. Therefore we refer to directional tangent lines, not just \textit{the} tangent line.
}
\printproblems
\input{exercises/12_06_exset_01}
\input{exercises/12_06_exset_02}
\input{exercises/12_06_exset_03}
\input{exercises/12_06_exset_04}
\input{exercises/12_06_exset_05}