\printproblems

\exercise{Find the volume $V$ inside the paraboloid $z=x^2 + y^2$ for $0 \le z \le 4$.}{$8\pi$}

\exercise{Find the volume $V$ inside the cone $z=\sqrt{x^2 + y^2}$ for $0 \le z \le 3$.}{$9\pi$}

\exercise{Find the volume $V$ of the solid inside both $x^2 + y^2 + z^2 = 4$ and $x^2 + y^2 = 1$.}{$4\pi(8-3^{3/2})/3$}

\exercise{Find the volume $V$ inside both the sphere $x^2 + y^2 + z^2 = 1$ and the cone $z=\sqrt{x^2 + y^2}$.}{}

\exercise{Prove\label{pr_cyl_jac} \autoeqref{eqn:cylinderint}.}{}

\exercise{Prove\label{pr_sph_jac} \autoeqref{eqn:sphereint}.}{}

\exercise{Evaluate $\ds\iint_{R} \sin \left( \frac{x+y}2 \right)\,\cos \left( \frac{x-y}2 \right)\,dA$, where $R$  is the triangle with vertices $(0,0)$, $(2,0)$ and $(1,1)$. (\emph{Hint: Use the change of variables $u=(x+y)/2$, $v=(x-y)/2$.})}{$1-\sin2/2$}

\exercise{Find the volume of the solid bounded by $z=x^2 + y^2$ and $z^2 = 4(x^2 + y^2 )$.}{}

\exercise{Find the volume inside the elliptic cylinder $\frac{x^2}{a^2} + \frac{y^2}{b^2} = 1$ for $0\le z\le 2$.}{$2\pi ab$}

\exercise{Show that the volume inside the ellipsoid $\frac{x^2}{a^2} + \frac{y^2}{b^2} + \frac{z^2}{c^2} = 1$ is $\frac{4\pi abc}{3}$. (\emph{Hint: Use the change of variables $x=au$, $y=bv$, $z=cw$, then consider \autoref{exmp_volsphere}.})\index{ellipsoid}}{}

\exercise{Let $D$ be the region bounded by $x^2+y^2=1$, $z=1$, and $z=3$.  Evaluate 
\[\iiint_D\frac{\sqrt{x^2+y^2}}{z^2}\, dV.\]}{$4\pi/9$}

\exercise{Let $D$ be the region inside both the sphere $x^2+y^2+z^2=1$ and the cone $z=\sqrt{x^2+y^2}$.  Evaluate the following.
\begin{enumext}[start=1]
\item $\ds\iiint_D z\,dV$
\item $\ds\iiint_D \frac1{1+x^2+y^2+z^2}\,dV$
\end{enumext}}{(a) $\pi/8$, (b) $\pi(2-\sqrt2)(4-\pi)/4$}

\exercise{Let $0<h<a$. Find the volume of the region bounded by the sphere $x^2+y^2+z^2=a^2$, $z=h$, and $z=a$.}{$\frac{2\pi}3(a^3-h^3)$}


\exercise{Find the mass of a (right circular) cylindrical solid of radius $R$ and height $h$ if the density at each point is proportional to its distance from the axis of the cylinder.}{$2\pi\alpha hR^3/3$ units of mass, where $\alpha$ is the proportionality constant}

\exercise{Find the mass of a (right circular) cylindrical solid of radius $R$ and height $h$ if the density at each point is proportional to the square of its distance from the base of the cylinder.}{$\pi\alpha h^3R^2/3$ units of mass, where $\alpha$ is the proportionality constant}

\exercise{Find the volume inside the upper nappe of the cone $z^2=x^2+y^2$ and below the plane $z=1$ using (a) cylindrical coordinates, (b) spherical coordinates.}{$\pi/3$ cubic units}

\exercise{Find the volume of the region bounded by the paraboloid $z=x^2+y^2$, the $xy$-plane, and the cylinder $x^2+y^2=1$.}{$\pi/2$ cubic units}

\exercise{Find the volume of the region inside the surface $\rho=1-\cos\phi$.}{$8\pi/3$ cubic units}

\exercise{Find the volume inside the sphere $x^2+y^2+z^2=1$ and below the plane $z=1/2$.}{$27\pi/24$ cubic units.}

\exercise{Find the volume inside the cylinder $x^2+y^2=1$ between the planes $z=0$ and $y+2z=1$.}{$\pi/2$ cubic units.}

\exercise{Use cylindrical coordinates to show that a right circular cone with base radius $R$ and height $h$ has volume $V=\pi R^2h/3$ cubic units.}{}

\exercise{Suppose a solid (right circular) cone with base radius $R$ and height $h$ is such that the density at each point is proportional to its distance from the axis of the cone. Find the mass.}{$\pi\alpha R^3h/6$ units of mass, where $\alpha$ is the proportionality constant.}

\exercise{Find the volume of the region bounded by $z=0$, $x^2+y^2=1$, and $z=\abs x+\abs y+1$.}{$\frac83+\pi$ cubic units}

\exercise{Evaluate $\ds\iiint_D z(x^2+y^2+z^2)^{-3/2}\,dV$, where $D$ is the region bounded below by $z=2$ and above by $x^2+y^2+z^2=16$.}{$\pi$}

\exercise{Evaluate $\ds\iiint_D y\,dV$ where $D$ is the region $x^2+y^2+z^2\le1$, below the first quadrant.}{$\pi/16$}

\exercise{Find the mass of a solid ball of radius $1$ if the density at points that are $\rho$ units from the center is $(1+\rho^2)^{-1}$.}{$4\pi-\pi^2$}

\exercise{Evaluate $\ds\iiint_D z\,dV$ where $D$ is the lens shaped region between $z=x^2+y^2$ and $z=1-x^2-y^2$.}{$\pi/8$}
 