\printconcepts

\exercise{What are the differences between the Divergence Theorems of \autoref{sec:greensthm} and this section?}{Answers will vary; in \autoref{sec:greensthm}, the Divergence Theorem connects outward flux over a closed curve in the plane to the divergence of the vector field, whereas in this section the Divergence Theorem connects outward flux over a closed surface in space to the divergence of the vector field.}

\exercise{What property of a vector field does the Divergence Theorem relate to flux?}{Divergence.}

\exercise{What property of a vector field does Stokes' Theorem relate to circulation?}{Curl.}

\exercise{Stokes' Theorem is the spatial version of what other theorem?}{Green's Theorem.}

% 14.4#5
%\exercise{Let $\vec F$ be a vector field and let $C_1$ and $C_2$ be any nonintersecting paths except that each starts at point $A$ and ends at point $B$. If \underline{\hskip.5in}$=0$, then $\int_{C_1} \vec F\cdot \vec T\ ds = \int_{C_2} \vec F\cdot \vec T\ ds$.}{$\curl \vec F$}
%
% 14.4#6
%\exercise{Let $\vec F$ be a vector field and let $C_1$ and $C_2$ be any nonintersecting paths except that each starts at point $A$ and ends at point $B$. If \underline{\hskip.5in}$=0$, then $\int_{C_1} \vec F\cdot \vec n\ ds = \int_{C_2} \vec F\cdot \vec n\ ds$.}{$\divv \vec F$}

\printproblems

\input{exercises/14-07-exset-01}

\input{exercises/14-07-exset-02}

\input{exercises/14-07-exset-03}

\input{exercises/14-07-exset-04}

\input{exercises/14-07-exset-05}

% 14.3#21
%\exercise{Prove part of \autoref{thm:conservative_field_curl}: let $\vec F =\bracket{M,N,P}$ be a conservative vector field. Show that $\curl \vec F = 0$.}{Since $\vec F$ is conservative, it is the gradient of some potential function. That is, $\nabla f = \bracket{f_x,f_y,f_z}= \vec F =\bracket{M, N, P}$. In particular, $M = f_x$, $N = f_y$ and $P=f_z$.
%
%Note that $\curl \vec F =\bracket{P_y - N_z, M_z-P_x, N_x-M_y}=\bracket{f_{zy} - f_{yz}, f_{xz} - f_{zx}, f_{yx} - f_{xy}}$, which, by \autoref{thm:mixed_partial}, is $\bracket{0,0,0}$.}

\exercise{Construct a M\"{o}bius strip from a piece of paper, then draw a line down its center (like the dotted line in \autoref{fig:mobius}(b)). Cut the M\"{o}bius strip along that center line completely around the strip. How many surfaces does this result in? How would you describe them? Are they orientable?}{}

\exercise{Use a computer algebra system to plot the M\"{o}bius strip parametrized as $\vecr(u,v) =$
  \[
   \bracket{\cos u \,(1+v\cos \tfrac{u}{2}),\sin u \,(1+v\cos \tfrac{u}{2}),v\sin \tfrac{u}{2}},
  \]
  where $0\le u\le2\pi$, $-\tfrac12\le v\le\tfrac12$ }{}

% todo solution to 15.7#33,34
\exercise{Let $\surfaceS$ be a closed surface and $\vec F$ a smooth vector field. Show that $\ds\iint_{\surfaceS} (\curl{\vec F}\,)\cdot\vecn\dd S= 0$. (\emph{Hint: Split $\surfaceS$ in half.})}{}

\exercise{Show that Green's Theorem is a special case of Stokes' Theorem. (Hint: Think of how a vector field $f(x,y)=P(x,y)i+Q(x,y)j$ in $\mathbb{R}^{2}$ can be extended in a natural way to be a vector field in $\mathbb{R}^{3}$.)}{}
