\printproblems

\exerciseset{In Exercises}{, use Green's Theorem to evaluate the given line integral around the curve $C$, traversed counterclockwise.}{

\exercise{$\ds\oint_C (x^2 - y^2 )\,dx + 2xy\,dy$; $C$ is the boundary of $R = \lbrace\,(x,y): 0 \le x \le 1,2x^2 \le y \le 2x \,\rbrace$}{$16/15$}

\exercise{$\ds\oint_C x^2 y\,dx + 2xy\,dy$; $C$ is the boundary of $R = \lbrace\,(x,y): 0 \le x \le 1,x^2 \le y \le x \,\rbrace$}{}

\exercise{$\ds\oint_C 2y\,dx - 3x\,dy$; $C$ is the circle $x^2 + y^2 = 1$}{$-5\pi$}

\exercise{$\ds\oint_C (e^{x^2} + y^2 )\,dx + (e^{y^2} + x^2 )\,dy$; $C$ is the boundary of the triangle with vertices $(0,0)$, $(4,0)$ and $(0,4)$}{}

\exercise{$\ds\oint_C\bracket{xy,y}\cdot d\vecr$; $C$ is the square with vertices $(0,0)$, $(1,0)$, $(1,1)$, $(0,1)$.}{$-1/2$}

\exercise{$\ds\oint_C\bracket{y^2,x^2}\cdot d\vecr$; $C$ is the triangle with vertices $(0,0)$, $(1,0)$, $(1,1)$.}{$1/3$}

\exercise{$\ds\oint_C\bracket{y\sin x,xy}\cdot d\vecr$; $C$ is the boundary of the region bordered by $\cos x$ and $\sin x$ for $\frac\pi4\le x\le\frac{5\pi}4$.}{$-\pi/2$}

}

\exercise{For $\vecf=\bracket{y^2,x^2}$, evaluate $\ds\oint_C\vecf\cdot d\vecr$ where $C$ is the boundary of the half disk of radius $1$ centered at the origin, in the first and second quadrants, traversed \emph{clockwise}.}{$4/3$}

\exercise{Repeat the previous problem, but use the unit half-disk (centered at the origin) in the fourth and first quadrants.}{$-4/3$}

\exercise{For $\vecf=\bracket{y^3,\sin y-x^3}$, evaluate $\ds\oint_C\vecf\cdot d\vecr$ where $C$ is the boundary of the annular region between $x^2+y^2=1/4$ and $x^2+y^2=1$, traversed so that the region is always on the left.}{$-45\pi/32$}


\exercise{Is there a potential $F(x,y)$ for $\vecf(x,y) = (y^2 + 3x^2 )\,\veci + 2xy\,\vecj$? If so, find one.}{Yes. $F(x,y)=xy^2+x^3$}

\exercise{Is there a potential $F(x,y)$ for $\vecf(x,y) = (x^3 \cos (xy) + 2x \sin (xy))\,\veci + x^2 y \cos (xy)\,\vecj$? If so, find one.}{}

\exercise{Is there a potential $F(x,y)$ for $\vecf(x,y) = (8xy+3)\,\veci + 4(x^2 + y)\,\vecj$? If so, find one.}{Yes. $F(x,y)=4x^2y+2y^2+3x$}

\exercise{Show that for any constants $a$, $b$ and any closed simple curve $C$, $\ds\oint_C a\,dx + b\,dy = 0$.}{}

\exercise{For\label{const_closed_int} the vector field \vecf\ as in \autoref{exmp_greenhole}, show directly that $\oint_{C}\vecf\cdot d\vecr = 0$, where $C$ is the boundary of the annulus $R =\{\,(x,y): 1/4 \le x^2 + y^2 \le 1\,\}$ traversed so that $R$ is always on the left.}{}

\exercise{Evaluate $\ds\oint_C  e^x \sin y\,dx + (y^3 + e^x \cos y)\,dy$, where $C$ is the boundary of the rectangle with vertices $(1,-1)$, $(1,1)$, $(-1,1)$ and $(-1,-1)$, traversed counterclockwise.}{}

\exercise{For a region $R$ bounded by a simple closed curve $C$, show that the area $A$ of $R$ is
  \[A = -\oint_C y\,dx = \oint_C x\,dy = \frac{1}{2}\oint_C x\,dy - y\,dx ,\]
  where $C$ is traversed so that $R$ is always on the left. (\emph{Hint: Use Green's Theorem and the fact that $A = \iint_{R} 1\,dA$}.)}{}

