\printproblems

\exerciseset{In Exercises}{, calculate $\int_C f(x,y)\,ds$ for the given function $f(x,y)$ and curve $C$.}{

\exercise{$f(x,y)=xy$; $\quad C: x=\cos t$, $y=\sin t$, $0 \le t \le \pi/2$}{$1/2$}

\exercise{$f(x,y)=\dfrac{x}{x^2 + 1}$; $\quad C: x=t$, $y=0$, $0 \le t \le 1$}{}

\exercise{$f(x,y)=2x+y$; $\quad C$: polygonal path from $(0,0)$ to $(3,0)$ to $(3,2)$}{$23$}

% todo find the answer to 15.1#4
\exercise{$f(x,y)=x+y^2$; $\quad C$: path from $(2,0)$ counterclockwise along the circle $x^2 + y^2 = 4$ to the point $(-2,0)$ and then back to $(2,0)$ along the $x$-axis}{}

\exercise{$f(x,y)=\sqrt{1+9xy}$; $\quad C$: the curve $y=x^3$ for $0\le x\le1$.}{$14/5$}

}

\exercise{A thin wire in the shape of the parbola $y=x^2$ for $1\le x\le4$ has mass density $\delta(x,y)=y/x$ g/cm. Find the total mass.}{$(65^{3/2}-5^{3/2})/12$ g}

\exercise{Use a line integral to find the lateral surface area of the part of the cylinder $x^2 + y^2 = 4$ below the plane $x+2y+z=6$ and above the $xy$-plane.}{$24\pi$}

\exerciseset{In Exercises}{, calculate $\int_C \vecf\cdot d\vecr$ for the given vector field $\vecf(x,y)$ and curve $C$.}{

\exercise{$\vecf(x,y) = \veci - \vecj$; $\quad C: x=3t$, $y=2t$, $0 \le t \le 1$}{}

\exercise{$\vecf(x,y) = y\,\veci - x\,\vecj$; $\quad C: x=\cos t$, $y=\sin t$, $0 \le t \le 2\pi$}{$-2\pi$}

\exercise{$\vecf(x,y) = x\,\veci + y\,\vecj$; $\quad C: x=\cos t$, $ y=\sin t$, $0 \le t \le 2\pi$}{}

\exercise{$\vecf(x,y) = (x^2 - y)\,\veci + (x-y^2 )\,\vecj$; $\quad C: x=\cos t$, $y=\sin t$, $0 \le t \le 2\pi$}{$2\pi$}

% todo find the answer to 15.1#12
\exercise{$\vecf(x,y)=xy^2 \,\veci + xy^3 \,\vecj$; $\quad C:$ the polygonal path from $(0,0)$ to $(1,0)$ to $(0,1)$ to $(0,0)$}{}

\exercise{$\vecf(x,y)=(x^2 + y^2 )\,\veci$; $\quad C: x=2+\cos t$, $y=\sin t$, $0 \le t \le 2\pi$}{$0$}

\exercise{$\vecf(x,y)=\bracket{x^2,xy}$; $\quad C$: the triangle with vertices $(1,0)$, $(0,1)$, $(-1,0)$ traversed counterclockwise.}{$1/3$}

\exercise{$\vecf(x,y)=\bracket{y^2,x^2}$; $\quad C$: the curve $y=1/x$ for $1\le x\le2$, traversed (a) left to right; (b) right to left.}{$-1/2$; $1/2$}

}

\exercise{Find the work done by the force $\vecf=\bracket{x,xy}$ moving an object along the curve $C$ where $C$ is the top half of the circle $x^2+y^2=1$ traversed from $(1,0)$ to $(-1,0)$.}{$2/3$ (units of work)}

\exercise{Find the work done by the force $\vecf=\bracket{x,xy}$ moving an object along the curve $C$ where $C$ is the the parabola $y=1-x^2$ traversed from (1, 0) to (-1, 0).}{$8/15$ (units of work)}

\exercise{Verify\label{pr_double_circ} that the value of the line integral in \autoref{exmp_lineintcyl} is unchanged when using the parametrization of the circle $C$ given in \autoeqref{eqn:lineintcylcwise}.}{}

\exercise{Show that if $\vecf \perp \vecr\primeskip'(t)$ at each point $\vecr(t)$ along a smooth curve $C$, then $\int_C \vecf\cdot d\vecr = 0$.}{}

\exercise{Show that if $\vecf$ points in the same direction as $\vecr\primeskip'(t)$ at each point $\vecr(t)$ along a smooth curve $C$, then $\int_C \vecf\cdot d\vecr = \int_C \norm{\vecf}\,ds$.}{}

\exercise{Prove\label{pr_rev_int} that $\int_C f(x,y)\,ds = \int_{-C} f(x,y)\,ds$. (\emph{Hint: Use \autoeqref{eqn:reversec}.})}{}

\exercise{Let $C$ be a smooth curve with arc length $L$, and suppose that $\vecf(x,y) = P(x,y)\,\veci + Q(x,y)\,\vecj$ is a vector field such that $\norm{\vecf(x,y)} \le M$ \smallskip for all $(x,y)$ on $C$. Show that\\
$\abs{\int_C \vecf\cdot d\vecr} \le ML$. (\emph{Hint: Recall that $\abs{\int_a^b g(x)\,dx} \le \int_a^b \abs{g(x)}\,dx$ for Riemann integrals.})}{}

\exercise{Prove that the Riemann integral $\int_a^b f(x)\,dx$ is a special case of a line integral.}{}
